---
title: "Challenge_1: Data Import, Description, and Transformation(1)"
author: "Jeff Scheminger"
description: ""
date: "09/20/2023"
format:
  html:
    df-print: paged
    css: "styles.css"
categories:
  - weekly_challenges
  - challenge_1
---

## Setup

If you have not installed the following packages, please install before loading them.

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(haven) #for loading other datafiles (SAS, STATA, SPSS, etc.)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

This first weekly challenge aims to practice the following skill sets: 1. Read datasets in different file types; 2. Describe the datasets; 3. Exploring a few basic functions of data transformation and wrangling; and 4. Present some descriptive statistics.

There will be coding components (reading datasets and data transformation) and writing components (describing the datasets and some statistical information). Please read the instructions for each part and complete your challenges.

## Creating your own R quarto project and make sure your working directory

(Will be demonstrated on Sep 20 and 21 lab meetings)

## Datasets

Please download the following dataset files from Canvas or Google Classroom and save them to a folder within your project working directory (i.e.: "yourworkingdiectory_data").

-   babynames.csv (Required) ⭐
-   ESS_5.dta (Option 1) ⭐ Codebook:
-   p5v2018.sav (Option 2)⭐ Codebook:
-   StateCounty2012.xlsx (Required)⭐⭐

Find the `_data` folder, then use the correct R command to read the datasets.

## Baby Names Dataset (Required)

1.  **Read the dataset "babynames.csv":**

```{r}
#Type your code here
setwd(Users/jscheminger/Library/CloudStorage/OneDrive-Personal/Documents/DACSS_Masters/DACSS_601_Fall_23)
```

2.  **Data Description: Please use the necessary commands and codes and briefly describe this data with a short writing paragraph answering the following questions.**

    ```{r}
    #Type your code here; and write a paragraph answering the questions.
    ```

    \(1\) What is the dimension of the data (# of rows and columns)?

    \(2\) What do the rows and columns mean in this data?

    \(3\) What is the unit of observation? In other words, what does each case mean in this data?

    \(4\) According to the lecture, is this a "tidy" data?

3.  **Data Transformation: use necessary commands and codes and answer the following questions.**

    ```{r}
    #Type your code here; and write a paragraph answering the questions.
    ```

    \(1\) How many unique male names, unique female names, and total unique names are in the data?

    \(2\) How many years of names does this data record?

    \(3\) Summarize the min, mean, median, and max of "Occurrence". (Must use summarize())

    \(4\) (Optional) Summarize the min, mean, median, and max of "Occurrence" by decade.

## 

**In the following part, choose either one of the two datasets to complete the tasks.**

## Optional 1: European Social Survey Dataset

1.  **Read the dataset "ESS_5.dta".**

    ```{r}
    #Type your code here
    ```

<!-- -->

2.  **Data Description: Please use the necessary commands and codes and briefly describe this data with a short writing paragraph answering the following questions.**

    \(1\) What is the dimension of the data (# of rows and columns)?

    ```{r}
    #Type your code here; and write a paragraph answering the questions.
    ```

    As we can see, this data is very large. We don't want to study the whole data. Let's just reload the following selected columns: idno, essroud, male, age, edu, income_10, eth_major, media (a standardized measure of the frequency of media consumption), and cntry.

    ```{r}
    #Type your code here; and write a paragraph answering the questions.
    ```

    \(2\) For the reloaded data, what do the rows and columns mean in this data?

    \(3\) What is the unit of observation? In other words, what does each case mean in this data?

    \(4\) According to the lecture, is this a "tidy" data?

3.  **Data Transformation: use necessary commands and codes and answer the following questions.**

    ```{r}
    #Type your code here; and write a paragraph answering the questions.
    ```

    \(1\) How many unique countries are in the data?

    \(2\) What is the range and average of the following variables: "age", "edu", and "media"? Must use summarize().

    \(3\) How many missing data (NA) are in the following variables: "eth_major" and "income_10"? (tips: use is.na())

## Optional 2: Polity V Data

1.  **Read the dataset "p5v2018.sav".**

    ```{r}
    #Type your code here
    ```

<!-- -->

2.  **Data Description: Please use the necessary commands and codes and briefly describe this data with a short writing paragraph answering the following questions.**

    ```{r}
    #Type your code here; and write a paragraph answering the questions.
    ```

    \(1\) What is the dimension of the data (# of rows and columns)?

    As we can see, this data contains many columns. We don't want to study the whole data. Let's keep the first seven columns and the eighth and ninth columns.

    ```{r}
    #Type your code here; and write a paragraph answering the questions.
    ```

    \(2\) For the reloaded data, what do the rows mean in this data? What do the columns (#2-#8) mean?

    \(3\) What is the unit of observation? In other words, what does each case mean in this data?

    \(4\) According to the lecture, is this a "tidy" data?

3.  **Data Transformation: use necessary commands and codes and answer the following questions.**

    ```{r}
    #Type your code here; and write a paragraph answering the questions.
    ```

    \(1\) How many unique countries are in the data?

    \(2\) How many years does this data record?

    Noted that in this data, missing data is coded as negative integers (-88, -77, and -66).

    \(3\) What is the range and average of the following variables: "democ" and "autoc"?

    \(4\) How many missing data (NA) in the following variables: "democ" and "autoc"? (tips: use is.na())

## Railraod Employee Data

1.  **Read the dataset "railroads.xls".**

    Many government organizations still use Excel spreadsheet to store data. This railraod dataset, publisehd by the Railroad Retirement Board, is a typical example. It records the number of employees in each county and state in 2012.

    **Please load the data in R in a clean manner. You can start with doing the following things step by step.**

    \(1\) Read the first sheet of the excel file;

    \(2\) Skipping the title rows;

    \(3\) Removing empty columns

    \(4\) Filtering "total" rows

    \(5\) Remove the table notes (the last two rows)

    ```{r}
    #Type your code here
    ```

<!-- -->

2.  **Data Description: Please use the necessary commands and codes and briefly describe this data with a short writing paragraph answering the following questions.**

    ```{r}
    #Type your code here; and write a paragraph answering the questions.
    ```

    \(1\) What is the dimension of the data (# of rows and columns)?

    \(2\) What do the rows and columns mean?

    \(3\) What is the unit of observation? In other words, what does each case mean in this data?

    \(4\) According to the lecture, is this a "tidy" data?

3.  **Data Transformation: use necessary commands and codes and answer the following questions.**

    ```{r}
    #Type your code here; and write a paragraph answering the questions.
    ```

    \(1\) How many unique counties and states are in the data? (tips: you can try using the across() function to do an operation on two columns at the same time)

    \(2\) What is the total number of employees (total_employees) in this data?

    \(3\) What is the min, max, mean, and median of "total_employees"

    \(4\) Which states have the most employees? And which countries have the most employees? (tips: use group_by() and arrange())
